SULI Report!

### 2015-08-20 doc/suli-report-swe.tex
Updated appendix A with the visualizing correlation plots. #81

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{mathtools}
\usepackage{caption}
\usepackage{subcaption}
%\usepackage{multicol}
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\author{Spencer Everett}
 \affiliation{Department of Physics, DePaul University.}%Lines break automatically or can be forced with \\
 \email{spencerweverett@gmail.com}

@@

One of the most successful probes of dark matter has been gravitational lensing. The path of light from distant `background' galaxies is bent when traveling through regions of space containing large amounts of `foreground' mass. The process is described by Einstein's general theory of relativity, but simply put -- the more mass, the more light bends. Light from different origins in a background galaxy are subject to different bending which results in a distortion of the galaxy image. As the foreground mass is known to be largely dark matter, gravitational lensing supplies a direct constraint on the mass of dark matter in that region.

@@

Using the physical separation distance and halo mass, the shear and convergence contribution of a single foreground halo is calculated using methods described in Wright and Brainerd \cite{lensing_calc}. The total convergence and shear at the center of the lightcone is simply the sum of the convergence and shear contributions of each halo contained in the lightcone. The \texttt{Pangloss} predicted lensed background ellipticity is then calculated using Equation \eqref{lensing}.

@@

Instead of comparing the ray-traced and \texttt{Pangloss} predicted ellipticities for individual galaxies, the lensing is characterized globally with correlation functions. The ellipticity-ellipticity correlation function measures how correlated the ellipticities of pairs of galaxies are as a function of separation distance, while the galaxy-mass correlation function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance. Both correlation functions are used in this work to measure how well the \texttt{Pangloss} framework models weak lensing by dark matter structures using the publicly available \texttt{TreeCorr} module written by Mike Jarvis\footnote{https://github.com/rmjarvis/TreeCorr}. Note that the ellipticity-ellipticity correlation function definition used in \texttt{TreeCorr} is slightly different than that used in most of the literature; for a derivation of the connection between Jarvis's definition and the more common Schneider definition \cite{schneider}, see the appendix.
%For readers that are unfamiliar with correlation functions in the context of cosmology or want a visual aid, see Appendix A.

@@

The first test of the halo model was with the ellipticity-ellipticity correlation function with a lightcone radius of 2 arcminutes which is given in Figure \ref{gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between 0.1 and 2 arcminutes. The correlation function of the intrinsic ellipticities plotted in blue shows no significant deviation from zero as expected, as before lensing the galaxies had random orientations. Both the ray-traced, plotted in green, and halo model, plotted in purple, ellipticities show positive correlation on small scales up to a separation of 1 arcminute. However, the ray-traced correlation is positive on all calculated scales while the halo model does not predict any significant corrlelation at separations above 1arcminute. Additionally, while the shapes of the correlation functions for both sets of lensed ellipticities aggree well until 1 arcminute, there is a clear systematic underprediction of shear power by the halo model on all scales. On average, the shear power is underpredicted by 117.0 percent.

@@

\subsection*{Galaxy-mass Correlation Function}

@@

\subsection*{Issues with the \texttt{Pangloss} Framework}

@@

\subsection*{Scaling Up \texttt{Pangloss}}

@@

The first statistic used in this work is the ellipticity-ellipticity ($\varepsilon-\varepsilon$) correlation function, which has two tangential components $\xi_\pm$ and a cross component $\xi_\times$. The components most interesting for gravitational lensing are the positive tangential component $\xi_+$, expected to be positive for lensed galaxies, and the cross component $\xi_\times$, which is expected to be zero. To get an intuition for what these quantities measure, first consider the correlation values for individual pairs of galaxies as shown in Figure \ref{corr_colormaps}. In each plot, the value of the correlation between two galaxy orientations is plotted as a colormap with a few sampled orientations plotted on top of the graph.

\begin{itemize}
\item Use as a brief introduction to the shear-shear and galaxy-mass correlation function as used in this work
    \begin{itemize}
    \item reference other (much more detailed) material for readers to find
    \end{itemize}
\item Include correlation value colormaps for $\xi_+$, $\xi_-$, and $\xi_\times$ along with discussion of correlation pattern
\item Include color-coded correlation visuals along with setup
    \begin{itemize}
    \item Galaxies around lens before lensing - correlation values scatter plot with no signal
    \item Galaxies around lens after lensing - correlation values scatter plot with signal
    \end{itemize}
\item B-modes and E-modes plot??
\end{itemize}

@@

Maybe do importance and/or lookup table after paper is submitted? Talk about in discussion?

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{mathtools}
\usepackage[format=plain]{caption}
\usepackage{subcaption}
%\usepackage{multicol}
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\author{Spencer Everett}
 \affiliation{Department of Physics, DePaul University,\\SLAC National Laboratory}
 %\affiliation{SLAC National Laboratory}
 \email{spencerweverett@gmail.com}

@@

One of the most successful probes of dark matter has been gravitational lensing. The path of light from distant `background' galaxies is bent when traveling through regions of space containing large amounts of `foreground' mass. Light from different origins in a background galaxy are subject to different bending which results in a distortion of the galaxy image. As the foreground mass is known to be largely dark matter, gravitational lensing supplies a direct constraint on the mass of dark matter in that region.

@@

Using the physical separation distance and halo mass, the shear and convergence contribution of a single foreground halo is calculated using methods described in Wright and Brainerd \cite{lensing_calc}. The total convergence and shear at the center of the lightcone is simply the sum of the convergence and shear contributions of each halo contained in the lightcone. The \texttt{Pangloss} predicted lensed source ellipticity is then calculated using Equation \eqref{lensing}.

@@

Instead of comparing the ray-traced and \texttt{Pangloss} predicted ellipticities for individual galaxies, the lensing is characterized globally with correlation functions. The ellipticity-ellipticity correlation function measures how correlated the ellipticities of pairs of galaxies are as a function of separation distance, while the galaxy-mass correlation function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance. For readers that are unfamiliar with correlation functions in the context of cosmology or want a visual aid, see appendix A. Both correlation functions are used in this work to measure how well the \texttt{Pangloss} framework models weak lensing by dark matter structures using the publicly available \texttt{TreeCorr} module written by Mike Jarvis\footnote{https://github.com/rmjarvis/TreeCorr}. Note that the ellipticity-ellipticity correlation function definition used in \texttt{TreeCorr} is slightly different than that used in most of the literature; for a derivation of the connection between Jarvis's definition and the more common Schneider definition \cite{schneider}, see appendix B.

@@

The first test of the halo model was with the ellipticity-ellipticity correlation function with a lightcone radius of 2 arcminutes which is given in Figure \ref{gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between 0.1 and 2 arcminutes. The correlation function of the intrinsic ellipticities plotted in blue shows no significant deviation from zero as expected, as before lensing the galaxies had random orientations. Both the ray-traced, plotted in green, and halo model, plotted in purple, ellipticities show positive correlation on small scales up to a separation of 1 arcminute. However, the ray-traced correlation is positive on all calculated scales while the halo model does not predict any significant corrlelation at separations above 1arcminute. Additionally, while the shapes of the correlation functions for both sets of lensed ellipticities aggree well until 1 arcminute, there is a clear systematic underprediction of shear power by the halo model on all scales. On average, the shear power is underpredicted by 117.0\%.

@@

\subsection*{Galaxy-Mass Correlation Function}

@@

\subsection*{Issues with the Pangloss Framework}

@@

\subsection*{Scaling Up Pangloss}

@@

The main statistic used in this work is the ellipticity-ellipticity ($\varepsilon-\varepsilon$) correlation function, which has two tangential components $\xi_\pm$ and a cross component $\xi_\times$. The components most interesting for gravitational lensing are the positive tangential component $\xi_+$, expected to be positive for lensed galaxies, and the cross component $\xi_\times$, which is expected to be zero. To get an intuition for what these quantities measure, first consider the correlation values for individual pairs of galaxies as shown in Figure \ref{corr_colormaps}. In each plot, the value of the correlation between two galaxies as a function of their orientation is plotted as a colormap with a few sampled galaxy orientations superimposed. For $\xi_+$, the correlation is positive for parallel galaxies, negative for perpendicular galaxies, and zero for galaxies offset by 45$^\circ$. In contrast, $\xi_\times$ is zero for all parallel and perpendicular orientations while nonzero for most 45$^\circ$ offsets.

%% Correlation Colormaps
\begin{figure*}
    \centering
    \begin{subfigure}[H]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/xip_colormap.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The positive tangential component of the $\varepsilon-\varepsilon$ correlation as a function of galaxy orientation plotted as a colormap. The correlation component is positive for parallel galaxies, negative for perpendicular galaxies, and zero for galaxies offset by 45$^\circ$.}
        \label{xip_colormap}
    \end{subfigure}
    ~
    %% Einstein Ring Figure
    \begin{subfigure}[H]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/xix_colormap.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The cross component of the $\varepsilon-\varepsilon$ correlation as a function of galaxy orientation plotted as a colormap. While the pattern is more complicated than that of the tangential component, note that the correlation value is zero for all preferred orientations after lensing.}
        \label{xix_colormap}
    \end{subfigure}
    \caption{Correlation pair colormaps for $\xi_+$ and $\xi_\times$.}
    \label{corr_colormaps}
\end{figure*}

Now consider

%% Correlation Colormaps
\begin{figure}[b!]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/corr_visualization_unlensed.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{A set of background sources around a foreground lens (blue dot). The size and orientation of each stick represents the intrinsic source ellipticity and is colored by the distance away from the lens.}
        \label{color_corr_unlensed}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/corr_dist_unlensed.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The distribution of correlation pairs from the source ellipticities in Figure \ref{color_corr_unlensed}. If the pair of galaxies are in the same color bin, then the correlation scatter point is colored in the same scheme. Otherwise, the correlation is not plotted.}
        \label{corr_dist_unlensed}
    \end{subfigure}

    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/corr_visualization_lensed.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The same set of background sources in Figure \ref{color_corr_unlensed} after (magnified for effect) lensing by the foreground mass. Regardless of initial orientation, the source galaxies have all at least partially aligned tangentially around the center lensing mass.}
        \label{color_corr_lensed}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/corr_dist_lensed.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The distribution of correlation pairs from the source ellipticities in Figure \ref{color_corr_lensed}. If the pair of galaxies are in the same color bin, then the correlation scatter point is colored in the same scheme. Otherwise, the correlation is not plotted.}
        \label{corr_dist_lensed}
    \end{subfigure}
    \caption{Correlation visualizaiton for unlensed galaxies.}
    \label{corr_colormaps}
\end{figure}

The $\varepsilon-\varepsilon$ correlation function is simply the average of all the correlations of each possible galaxy pair. Before lensing, galaxies should have uniformly distributed orientations as the universe is isotropic. Thus when averaged, both $\left<\xi_+\right>$ and $\left<\xi_\times\right>$ should be zero. However, as gravitational lensing tends to shear source images tangentially around a foreground mass,

@@

%% Appendix A
\begin{itemize}
\item Use as a brief introduction to the shear-shear and galaxy-mass correlation function as used in this work
    \begin{itemize}
    \item reference other (much more detailed) material for readers to find
    \end{itemize}
\item Include correlation value colormaps for $\xi_+$, $\xi_-$, and $\xi_\times$ along with discussion of correlation pattern
\item Include color-coded correlation visuals along with setup
    \begin{itemize}
    \item Galaxies around lens before lensing - correlation values scatter plot with no signal
    \item Galaxies around lens after lensing - correlation values scatter plot with signal
    \end{itemize}
\item B-modes and E-modes plot??
\end{itemize}
